Characterizing System Dynamics using Diffusion Maps: A Chemotaxis Case Study

### 2014-07-29 chemotaxis/prl_paper.tex
added small edits to chemotaxis paper

Many dynamical systems are governed by stochastic motions at the microscale which give rise to smooth, coherent dynamics at the macroscale.
%
The relationship between the microscopic and macroscopic dynamics is often not immediately obvious and either requires some modeling assumptions or extensive observation of the system.
%
We demonstrate how extracting a parameterization of microscale data can be accomplished automatically using manifold learning techniques, and that this parameterization is consistent with the governing macroscopic dynamical model. 
%
These data-driven algorithms need no {\em a priori} knowledge about the appropriate macroscopic model, but instead, only require the appropriate observers and pairwise metric between observations in order to recover the governing macroscopic variables.
%
We also show how these algorithms can detect changes in the macroscopic dynamical behavior as a function of the microscopic system parameters. 
%
We illustrate these ideas using a model which arises in studies of cellular chemotaxis.

@@

In this letter, we will present how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
%
Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison \cite{mallat2012group}. 
%
We will show that both are essential for data-driven algorithms to successfully uncover the underlying system dynamics.

We will illustrate these points using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}.
%
In this example, the macroscopic dynamics vary depending on the value of a single system parameter.
%
We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
%
This case study exhibits several key points which allow us to demonstrate the strength of our approach.
%
First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
%
Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
%
Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
%
We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between pairs of observations uncovers a description of the microscopic data which is consistent with the macroscopic model.
%
Furthermore, these data-driven methods allow us to detect changes in dynamical behavior resulting from changes in the system parameters.

@@

The microscopic model consists of a collection of $N$ particles whose states are defined by their positions and velocities. 
%
Let $x_i(t)$ and $v_i(t)$ denote the position and velocity of particle $i$ at time $t$, respectively.
%
The velocity of each particle is either $\pm s$, where $s$ is a (fixed) speed. 
%
We initialize the particles such that
\begin{equation}\label{eqn:system}
\begin{aligned}
x_i(0) & = 0 \\
\mathbb{P} \{ v_i(0) = +s \} & = p
\end{aligned}
\end{equation}
where $p$ is an initial probability.
%
The velocity of each particle randomly switches between $\pm s$ following an (independent) Poisson process with rate $\lambda$.
%
We would like to note that we have chosen a one-parameter family of initial conditions.
%
Other initial conditions could easily be chosen, but the simplicity of this setup will allow us to easily illustrate our main points.

This particular example has a known analytic macroscopic equation that governs the overall system behavior.
%
For a large collection of particles ($N \rightarrow \infty$), the system can be described by the probability density of the particles.
%
Let $\rho(x, t)$ denote the probability density of the particles, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the densities of the particles moving towards the positive and negative axis directions, respectively.
%
It can be shown that, as $N \rightarrow \infty$, the densities obey the following set of partial differential equations (PDEs) \cite{othmer2000diffusion}:
\begin{equation} \label{eqn:coupled_pdes}
\begin{aligned}
\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = -\lambda \rho^+ +\lambda \rho^- \\
\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = \lambda \rho^+ -\lambda \rho^- 
\end{aligned}
\end{equation}
%
Alternatively, \eqref{eqn:coupled_pdes} can be rewritten as one, second--order PDE:
\begin{equation} \label{eq:second_order_pde}
\frac{\partial^2 \rho}{\partial t^2} + 2 \lambda \frac{\partial \rho}{\partial t} = s^2 \frac{\partial ^2 \rho}{\partial x^2}
\end{equation}
%
%We assume that $s^2/\lambda = D$ is constant, so that the dynamics of the probability density are governed by a {\em single} parameter $\lambda$.
%
From \eqref{eq:second_order_pde}, we see that, for fixed values of $\lambda$ and $s$, the macroscopic state of the system (the probability density of the particles) is a function of two independent variables: $p$, which controls the initial distribution of the particles, and $t$, the time. 
%
This implies that for fixed $\lambda$ and $s$, the {\em microscopic} data in a high-dimensional ambient space (e.g., the positions of all $N$ particles) should lie on a two-dimensional manifold parameterized by $p$ and $t$, 
and uncovering the low-dimensional structure of the microscopic data reveals the governing macroscopic variables.

@@

For general dynamical systems, the governing macroscopic variables and asymptotic behavior may not be immediately obvious, and instead need to be inferred from the microscopic data.
%
However, similarly to the chemotaxis model, we expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
%
We use diffusion maps \cite{coifman2005geometric}, which is a manifold learning technique, to uncover such low-dimensional structure and to analyze the results of the microscopic chemotaxis simulations.
%
Compared to principal components analysis (PCA) \cite{shlens2005tutorial}, a standard linear technique, diffusion maps provides a parameterization of data which lie on a low-dimensional, (possibly) {\em nonlinear} structure in a high-dimensional space.
%
Diffusion maps is one of many recently introduced manifold learning techniques that could be used to analyzed this data \cite{roweis2000nonlinear, tenenbaum2000global, Belkin2003}.

@@

This paper aims to bridge the fields of data mining and dynamical systems. 
%
For data mining methods to be informative with regards to the underlying system dynamics, processing the appropriate statistical moments or observables instead of individual particles is essential. 
%
In addition, one must also define the appropriate distance metric between the observations.
%
These two components induce the ``correct" Riemannian geometry and allow for informative analyses, such as meaningful and compact parameterization, phase shift detection, and dynamical characterization, through manifold learning.
%
In this paper, these concepts were demonstrated using a dynamical model of cellular chemotaxis with a known macroscopic behavior.
%
We showed that, using histograms as observers and EMD as a distance metric, diffusion maps can uncover the underlying macroscopic dynamical behavior which is consistent with the analytic macroscopic model.
%
We are confident that such techniques can help inform modeling efforts for systems where the macroscopic dynamics and behavior are not currently known.

>>>>

Many dynamical systems are governed by stochastic motions at the microscale which give rise to smooth, coherent dynamics at the macroscale.
%
The relationship between the microscopic and macroscopic dynamics is often not immediately obvious and either requires some modeling assumptions or extensive observation of the system.
%
We demonstrate automatically extracting a parameterization of microscale data using manifold learning techniques, and show that this parameterization is consistent with the governing macroscopic dynamical model. 
%
These data-driven algorithms need no {\em a priori} knowledge about the appropriate macroscopic model, but instead, only require the appropriate observers and pairwise metric between observations.
%
We also show how these algorithms can detect changes in the macroscopic dynamical behavior as a function of the microscopic system parameters. 
%
We illustrate these ideas using a model which arises in studies of cellular chemotaxis.

@@

In this letter, we will demonstrate how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
%
Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison \cite{mallat2012group}. 
%
We will show that both are essential for data-driven algorithms to successfully uncover the underlying system dynamics.

We will illustrate these points using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}.
%
In this example, the macroscopic dynamics vary depending on the value of a single system parameter.
%
We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
%
This case study exhibits several key points which allow us to demonstrate the strength of our approach.
%
First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
%
Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
%
Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
%
We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between pairs of observations uncovers a parameterization of the microscopic data which is consistent with the macroscopic model.
%
Furthermore, these data-driven methods allow us to detect changes in dynamical behavior resulting from changes in the system parameters.

@@

The microscopic model consists of a collection of $N$ particles whose states are defined by their positions and velocities. 
%
Let $x_i(t)$ and $v_i(t)$ denote the position and velocity, respectively, of particle $i$ at time $t$.
%
The velocity of each particle is either $\pm s$, where $s$ is a (fixed) speed. 
%
We initialize the particles such that
\begin{equation}\label{eqn:system}
\begin{aligned}
x_i(0) & = 0 \\
\mathbb{P} \{ v_i(0) = +s \} & = p
\end{aligned}
\end{equation}
where $0 < p < 1$ is the probability of a particle initially moving to the right.
%
The velocity of each particle randomly switches between $\pm s$ following an (independent) Poisson process with rate $\lambda$.
%
We would like to note that we have chosen a very specific one-parameter family of initial conditions, which lead to simple dynamics and straightforward illustration of our main points.
%
More complex initial conditions could easily be chosen.

This particular example has a known analytic macroscopic equation that governs the overall system behavior.
%
For a large collection of particles ($N \rightarrow \infty$), the system can be described by the probability density of the particles.
%
Let $\rho(x, t)$ denote the probability density of the particles, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the densities of the particles moving towards the positive and negative axis directions, respectively.
%
It can be shown that, as $N \rightarrow \infty$, the densities obey the following set of partial differential equations (PDEs) \cite{othmer2000diffusion}:
\begin{equation} \label{eqn:coupled_pdes}
\begin{aligned}
\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = -\lambda \rho^+ +\lambda \rho^- \\
\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = \lambda \rho^+ -\lambda \rho^- 
\end{aligned}
\end{equation}
%
Alternatively, \eqref{eqn:coupled_pdes} can be rewritten as one, second--order PDE:
\begin{equation} \label{eq:second_order_pde}
\frac{\partial^2 \rho}{\partial t^2} + 2 \lambda \frac{\partial \rho}{\partial t} = s^2 \frac{\partial ^2 \rho}{\partial x^2}
\end{equation}
%
%We assume that $s^2/\lambda = D$ is constant, so that the dynamics of the probability density are governed by a {\em single} parameter $\lambda$.
%
From \eqref{eq:second_order_pde}, we see that, for fixed values of $\lambda$ and $s$, the macroscopic state of the system (the probability density of the particles) is a function of two independent variables: $p$, which controls the initial distribution of the particles, and $t$, the time. 
%
This implies that for fixed $\lambda$ and $s$, the {\em microscopic} data in a high-dimensional ambient space (e.g., the positions of all $N$ particles) should lie on a two-dimensional manifold parameterized by $p$ and $t$, 
and uncovering the low-dimensional structure of the microscopic data reveals this manifold.

@@

For general dynamical systems, the governing macroscopic variables and asymptotic behavior may not be immediately obvious, and instead need to be inferred from the microscopic data.
%
We expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
%
We use diffusion maps \cite{coifman2005geometric}, which is a manifold learning technique, to uncover such low-dimensional structure and to analyze the results of the microscopic chemotaxis simulations.
%
Compared to principal components analysis (PCA) \cite{shlens2005tutorial}, a standard linear technique, diffusion maps provides a parameterization of data which lie on a low-dimensional, (possibly) {\em nonlinear} structure in a high-dimensional space.
%
Diffusion maps is one of many recently introduced manifold learning techniques that could be used to analyzed this data \cite{roweis2000nonlinear, tenenbaum2000global, Belkin2003}.

@@

This paper aims to bridge the fields of data mining and dynamical systems. 
%
For data mining methods to be informative with regards to the underlying system dynamics, processing the appropriate statistical moments or observables instead of individual particles is essential. 
%
In addition, one must also define the appropriate distance metric between the observations.
%
These two components induce the ``correct" Riemannian geometry and allow for informative analyses, such as meaningful and compact parameterization, phase shift detection, and dynamical characterization, through manifold learning.
%
In this paper, these concepts were demonstrated using a dynamical model of cellular chemotaxis with a known macroscopic behavior.
%
We showed that, using histograms as observers and EMD as a distance metric, diffusion maps can uncover a parameterization of the microscale data which is consistent with the analytic macroscopic model.
%
We are confident that such techniques can help inform modeling efforts for systems where the macroscopic dynamics and behavior are not currently known.
